The effects of context inference on motor learning: savings, de-adaptation and spontaneous recovery

### 2021-08-30 adaptation/article/article.tex
Added abstract. Edited discussion a bit

\abstract{Abstract}
Very interesting abstract. Much wow.

@@

\subsection{Extended explanatory power}
Context is often directly observable, either via environmental cues or via feedback observed after actions. In many behavioral experiments, for example, the context of the current block of trials is easily identifiable by the color of a cue light \citep[e.g.][]{Ethier_Spontaneous_2008}, the posture of the hand \citep[e.g.][]{Gandolfo_Motor_1996}, the target of the movement \citep[e.g.][]{Lee_Dual_2009} or by the workspace in which the task is being performed \citep[e.g.][]{Shadmehr_Adaptive_1994}. The context can also be selected by the agent itself. For example, a person can decide the level of care devoted to moving an egg around, which determines the way in which the person would move their hands. In these cases, the context is clearly determined and known to the agent and switches between different adaptations are immediate.

In other cases, however, the context might not be directly observable, and context inference takes the form of an evidence-accumulating process that can take any amount of time to be sure of the context, including infinite. It is in these cases where the effects of context inference are most noticeable. Laboratory experiments with this property are common. While many experiments exist that give probabilistic cues [EGs], or cues that only reveal part of the context [EGs], evidence accumulation is not limited to these explicit cases. Indeed, as we argued in the Results section, many experiments inadvertently include partial contextual information used by participants.

The most direct secondary contextual information comes in the form of reward and end-point feedback. Participants are told whether they obtained the desired reward at the end of a trial and are shown the end point of their movement. When a participant observes an error that is too large to explain with the expected motor variability, this information can be used to infer that the inferred context might be wrong. This is the case of the experiments by \cite{Oh_Minimizing_2019} shown in \fref{fig:oh-2019}B-C: if the adaptation is high, changes in context produce errors much larger than those of motor variability, and a context switch is easily and immediately identified.; if adaptation is low, the errors produced by context switching are closer in magnitude to motor variability and evidence accumulation is necessary.

The same rationale explains the results from \cite{Davidson_Scaling_2004}: motor learning, which in our model is gated by context inference, is minimal for errors close to 2 and -2 (see their figure 2E). This is because an error of 2 or -2 signals that the participant incorrectly identified the context (as adaptation has a magnitude of 1). It is important to note that they did not calculate a learning of exactly zero in these near-maximal-error trials, which suggests that participants are doing \textit{post hoc} learning: after observing the outcome of their actions, they identify that they had inferred the context incorrectly, and can then adapt the movements for the correct context.

A subtler source of information can be found in long pauses between blocks of adaptation trials. \cite{Ethier_Spontaneous_2008} performed an experiment where the adaptation blocks were divided into sub-blocks of 60 trials each, with a long 30s pause between sub-blocks. The authors observed that at the beginning of each sub-block, participants showed an average dip in their learned adaptation, suggesting that they were partially returning to baseline (O) after each pause. This can be explained by context inference, as a long pause could prompt participants to infer that a switch had occured. Because this was inferred without any information that points at the identity of the new context, prior beliefs take over. This means that participants rely on their belief of the underlying probability of observing any of the known contexts, which is dominated by the previously observed context A, but now includes a component of the baseline O, as it is the most common in their lives.

Error-clamp (E) trials present another insight. During E trials, error is artificially kept at zero. This takes different forms, depending on the type of experiment. If error is kept at zero, one could assume that participants would continue doing what they were doing before, as there is no reason (no observed error) to inferr a change in context. However, this is almost never the case \cite[e.g.][]{Smith_Interacting_2006,Ethier_Spontaneous_2008,Forano_Timescales_2020,Vaswani_Decay_2013,Scheidt_Persistence_2000}. Instead, participants slowly reduce their adaptation, often displaying spontaneous recovery \cite[e.g.][]{Smith_Interacting_2006}, i.e. an initial response in the direction of one of the learned adaptations. Our model provides a principled account of this behavior: while the observed error (i.e. the cursor on the screen) is kept at zero, participants still receive proprioceptive feedback which tells them that something has changed. The natural variability in a participant's behavior will lead her to expect observing errors, which clashes with the observed zero error. This prompts the participant to re-evaluate their inferred context, which can partially activate a previously-learned adaptation, as we showed in \fref{fig:vaswani-2013}. This can be seen in \cite{Criscimagna-Hemminger_Consolidation_2008}, where the authors showed that introducing long periods before the E block begins lowers the initial force that participants exerted on the mechanical arm during the E block; longer periods of time make context inference revert to the prior expectation that a new baseline block begins, because participants are free to move their arm about during the pause \todo{check this}.

@@

[Maybe: talk about generalization in experiments with different workspaces]

@@

As we already discussed above, the inclusion of reliable sensory contextual cues (e.g. lights whose color uniquely identify a context) makes switching immediate, as in the experiments by \cite{Kim_Neural_2015}. We expect that the same effect would be observed in error-clamp trials. If the E block is learned by participants during training, it might still be difficult for them to identify that an E block has started, which would create delays similar to those in \fref{fig:vaswani-2013}. However, the model predicts that if a visual cue is introduced that identifies the E block, participants would immediately switch to their baseline behavior, no longer displaying spontaneous recovery, lag, nor the slow return to baseline. This immediate switch in the presence of contextual cues would persist even if endpoint feedback is manipulated as \cite{Vaswani_Decay_2013} did.

As we discussed in the Results section, the volatility of the environment (i.e. unpredictable switching between contexts) lowers the measured learning rate. Our model predicts that this effect would be greatly reduced if reliable contextual cues were introduced: if a participant can identify the context of the current trial before any decision or observation has occurred, the learning rate would not be affected by the volatility of the environment. This would manifest in \fref{fig:herzfeld-2014} in the form of no discernible differences between the two groups of participants.

@@

The model also predicts an effect reminescent of multisensory integration: in order to integrate contextual information from conflicting sources (e.g. probabilistic visual cues and noisy endpoint feedback), the weight placed on a source increases with its precision (the inverse of its variance). Such integration would manifest itself in experiments in which observations are noisy, as in the experiments by \cite{Kording_Bayesian_2004}, in which the position of the finger was obscured and instead participants are shown a blurry cursor which was some times shifted from its real position. If the added observation noise gives evidence for a particular context (the real underlying context or another one) and a visual cue gave partial information for another context, the participants' behavior would be more consistent with the most reliable source of contextual information.

>>>>

\abstract{Abstract}
Humans are able to adapt motor commands to changing dynamics of the environment and of the body itself. While we must adapt to slow changes in our body dynamics, e.g. because of muscle exhaustion throughout the day, we can also adapt to sudden changes in the environment. For example, when picking up a heavy objects, we must adapt our motor commands to be able to move the object around. The size and material of an object provide us with contextual information which we can use to prepare the correct response to moving the object, and our ability to switch between clear contexts (e.g. between lifting a big rock and small piece of paper) has been well studied. However, contextual information is not always clear or realiable (e.g. a big object of unknown material). In this work, we focus on the behavioral phenomena effected by imperfect contextual information and present a hierarchical computational model for motor adaptation based on exact Bayesian inference, which explicitly takes into account contextual information and how the dynamics of context inference affect learning and action selection. We show how our model explains the effects of sensory cues and propriosceptive information on switching between tasks, the effects of environmental volatility and predictability on learning, and the effects of previously-learned adaptations on switching between tasks, in addition to the well studied savings, de-adaptation and spontaneous recovery.

@@

\subsection{Extended explanatory power}
Context is often directly observable, either via environmental cues or via feedback observed after actions. In many behavioral experiments, for example, the context of the current block of trials is easily identifiable by the color of a cue light \citep[e.g.][]{Ethier_Spontaneous_2008}, the posture of the hand \citep[e.g.][]{Gandolfo_Motor_1996}, the target of the movement \citep[e.g.][]{Lee_Dual_2009} or by the workspace in which the task is being performed \citep[e.g.][]{Shadmehr_Adaptive_1994}. The context can also be selected by the agent itself. For example, a person can decide the level of care devoted to moving an egg or a lead ball around, which determines the way in which the person would move their hands. In these cases, the context is clearly determined and known to the agent and switches between different adaptations are immediate.

In other cases, however, the context might not be directly observable, and context inference takes the form of an evidence-accumulating process that can take any amount of time to be sure of the context. It is in these cases where the effects of context inference are most noticeable. While many experiments exist that give probabilistic contextual information \cite[e.g.][]{Scholz_uncontrolled_1999,Behrens_Learning_2007,Nassar_Dissociable_2019}, evidence accumulation is not limited to these explicit cases. Indeed, as we argued in the Results section, many experiments inadvertently include partial contextual information used by participants.

Contextual information can come in the form of reward and end-point feedback. Participants are told whether they obtained the desired reward at the end of a trial and are shown the end point of their movement. When a participant observes an error that is too large to explain with the expected motor variability, this information can be used to infer that the inferred context might be wrong. This is the case of the experiments by \cite{Oh_Minimizing_2019} shown in \fref{fig:oh-2019}B-C, as well as those by \cite{Zarahn_Explaining_2008}: if the adaptation is high, changes in context produce errors much larger than those of motor variability, and a context switch is easily and immediately identified; if adaptation is low, the errors produced by context switching are closer in magnitude to motor variability and evidence accumulation is necessary.

The same rationale explains the results from \cite{Davidson_Scaling_2004}: motor learning, which in our model is gated by context inference, is minimal for errors close to 2 and -2 (see their figure 2E). This is because an error of 2 or -2 signals that the participant incorrectly identified the context (as adaptation has a magnitude of 1).

A subtler source of information can be found in long pauses between blocks of adaptation trials. \cite{Ethier_Spontaneous_2008} performed an experiment where the adaptation blocks were divided into sub-blocks of 60 trials each, with a long 30s pause between sub-blocks. The authors observed that at the beginning of each sub-block, participants showed an average dip in their learned adaptation, suggesting that they were partially returning to baseline (O) after each pause. This can be explained by context inference, as a long pause could prompt participants to infer that a switch had occured. Because this was inferred without any information that points at the identity of the new context, prior beliefs take over. This means that participants rely on their belief of the underlying probability of observing any of the known contexts, which is dominated by the previously observed context A, but now includes a component of the baseline O, as it is the most common one in everyday life.

Error-clamp (E) trials present another insight. During E trials, error is artificially kept at zero. This takes different forms, depending on the type of experiment. If error is kept at zero, one could assume that participants would continue doing what they were doing before, as there is no reason (no observed error) to inferr a change in context. However, this is almost never the case \cite[e.g.][]{Smith_Interacting_2006,Ethier_Spontaneous_2008,Forano_Timescales_2020,Vaswani_Decay_2013,Scheidt_Persistence_2000,Pekny_Protection_2011}. Instead, participants slowly reduce their adaptation, often displaying spontaneous recovery \cite[e.g.][]{Smith_Interacting_2006}, i.e. an initial response in the direction of one of the learned adaptations. Our model provides a principled account of this behavior: while the observed error (i.e. the cursor on the screen) is kept at zero, participants still receive proprioceptive feedback which tells them that something has changed. The natural variability in a participant's behavior will lead her to expect observing errors, which clashes with the observed zero error. This prompts the participant to re-evaluate their inferred context, which can partially activate a previously-learned adaptation, as we showed in \fref{fig:vaswani-2013}. \cite{Pekny_Protection_2011} found similar results, demonstrating that the duration of the previously-observed adaptation block also affects behavior in the E block. Additionally, \cite{Criscimagna-Hemminger_Consolidation_2008} showed that introducing long periods before the E block begins lowers the initial force that participants exerted on the mechanical arm during the E block; longer periods of time make context inference revert to the prior expectation that a new baseline block begins, because participants are free to move their arm about during the pause.

@@

@@

\subsubsection{Error-clamp as a context}
The inclusion of reliable sensory contextual cues (e.g. lights whose color uniquely identify a context) makes switching immediate, as in the experiments by \cite{Kim_Neural_2015}. We expect that the same effect would be observed in error-clamp trials. If the E block is learned by participants during training, it might still be difficult for them to identify that an E block has started, which would create delays similar to those in \fref{fig:vaswani-2013}. However, the model predicts that if a visual cue is introduced that identifies the E block, participants would immediately switch to their baseline behavior, no longer displaying spontaneous recovery, lag, nor the slow return to baseline. This immediate switch in the presence of contextual cues would persist even if endpoint feedback is manipulated as \cite{Vaswani_Decay_2013} did.

\subsubsection{Volatility has no effect on learning}
The volatility of the environment (i.e. unpredictable switching between contexts) lowers the measured learning rate. Our model predicts that this effect would be greatly reduced if reliable contextual cues were introduced: if a participant can identify the context of the current trial before any decision or observation has occurred, the learning rate would not be affected by the volatility of the environment. This would manifest in \fref{fig:herzfeld-2014} in the form of no discernible differences between the two groups of participants.

@@

\subsubsection{Multi-source integration}
The model also predicts an effect reminescent of multisensory integration: in order to integrate contextual information from conflicting sources (e.g. probabilistic visual cues and noisy endpoint feedback), the weight placed on a source increases with its precision (the inverse of its variance). Such integration would manifest itself in experiments in which observations are noisy, as in the experiments by \cite{Kording_Bayesian_2004}, in which the position of the finger was obscured and instead participants are shown a blurry cursor which was some times shifted from its real position. If the added observation noise gives evidence for a particular context (the real underlying context or another one) and a visual cue gave partial information for another context, the participants' behavior would be more consistent with the most reliable source of contextual information.
